Think deep for reef resilience

### 2025-08-17 paper/paper_draft.qmd
Reshuffle methods section
@ConnectedSystems I have reshuffled methods paper section to have connectivity methods after introducing the model because we mention ADRIA-CoralBlox implementation of conn in that section before we have introduced the modelling platform.

Also removed results/discussion for DHW-tolerance. DHW-tolerance is now only mentioned in intro and supplementary methods

Depth characteristics for each reef were extracted using 10m resolution bathymetry data made
available in GBRMPA [-@GBRMPA2020]. For each reef, bathymetry data were spatially matched
with the 3806 reef outlines through geometric intersection analysis. Depth values within
each reef polygon were summarized using zonal statistics. All depth values were converted to
positive values (depths below sea level). The median depth was used to provide an indicative
depth value for each reef. Code, and the resulting depth estimates, are available at
**\[cite Canonical Reefs repo\]**.

While the bathymetry data adopted for the given assessment are said to be accurate up to 15m
from Mean Sea Level [@GBRMPA2020], there are many sources of uncertainty in the depth
estimates used for this study. These are outlined further in the Discussion. Notably, the
adopted depth estimates were derived from bathymetries based on satellite imagery. The
bathymetry data can be limited in their alignment with the established GBRMPA reef outlines
used to characterise reef locations and their areal extent. A Quality Assessment and Control
process identified reefs with potential data limitations: reefs with no overlapping
bathymetry data, reefs with minimum depth values above sea level (indicating potential data
artifacts), and reefs where bathymetry pixels covered less than 5% of the reef polygon area.
This process removed 842 reefs, or 22%, from the initial set of 3806 reefs. Furthermore,
reefs that are not associated with GBRMPA Management areas [@GBRMPA2007] or reefal
bioregions [@GBRMPA2024] were also removed from the investigation. Bioregions which
contained fewer than three reefs were also excluded. As a result, 1377 reefs were excluded
(36%), leaving a total of 2429 remaining for the purpose of this investigation.

@@

These GCMs were selected based on based on evaluation conducted by Grose et al.,
[-@grose2023], indicating their ability to recreate historical atmospheric temperature and
precipitation conditions in the Australasian region. They span a range of Equilibrium
Climate Sensitivity (ECS) values within the Intergovernmental Panel on Climate Change (IPCC)
assessed likely range. The IPCC defines ECS as the long term warming caused by a doubling of
carbon dioxide above its pre-industrial concentration [@forster2021]. While this assessment
does not consider oceanic data, Grose et al. [-@grose2023] provide a region-specific
selection mechanism which we use as an indicator of general performance of GCMs for the
region. An overview of the selected GCMs are provided in @tbl-gcms. Note here that Grose et
al., [-@grose2023] assessed GFDL-ESM4, whereas Dixon et al., [-@dixon2022; -@dixon2023]
provides data based on GFDL-CM4. Similarly, data for ACCESS-CM2 was used as a stand-in for
ACCESS-ESM1-5. These substitutions are assumed appropriate given they share the same
fundamental representation of the physical climate system.

Projections using Shared Socio-economic Pathway (SSP) 2-4.5 was used to represent the
plausible emission pathway. It represents a moderate emissions scenario with continued CO2
emissions comparable to present day until 2050, after which emissions gradually decline
[@hausfather2020; IPCC -@climate2023]. The selected GCMs represent 2.5°C to 4.31°C of global
atmospheric warming in the 2081-2100 period relative to 1850-1900. The use of this range of
GCMs therefore allow for a range of possible climate futures to be explored, capturing the
uncertainty in future responses to CO2 increases (@sfig-ecs; @fig-methods-dhws; [@grose2023;
@shepherd2018]) and enables the assessment of model projected coral reef behaviour over a
broad range of independently derived climate futures [@abramowitz2019].

@@

Degree Heating Weeks (DHWs) were then derived using the "HotSpot SST Anomaly" approach
[@skirving2020]. The "HotSpot SST Anomaly" represents the thermal stress experienced by
coral reefs and is calculated as the difference between daily SST values and the maximum of
a pixel's monthly mean climatology across all twelve months (i.e., the maximum monthly mean
SST). This approach uses the annual maximum temperature as the baseline because corals are
adapted to withstand their local warmest conditions under normal seasonal cycles. When
temperatures exceed this seasonal maximum, corals experience thermal stress that can lead to
bleaching. Only positive anomalies, greater than 1°C, are considered relevant for coral
stress, as temperatures below the seasonal maximum do not contribute to bleaching risk.

DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the 12-week
rolling sum of positive HotSpot SST Anomaly values that are greater than 1°C, expressed in
units of °C-weeks. The 12-week accumulation period reflects the cumulative nature of coral
thermal stress. Brief temperature spikes may be tolerable, but sustained or repeated
exposure to elevated temperatures causes progressively greater physiological stress and
bleaching severity. DHW values integrate both the magnitude and duration of thermal stress:
a moderate anomaly (e.g., 1°C) sustained for 12 weeks produces the same DHW value as a
severe anomaly (e.g., 3°C) lasting 4 weeks, though the biological impacts may differ. DHW
estimates for each indicative reef (see @GBRMPA2022) were then determined by taking the mean
DHW value for all pixels touching each reef outline. The process described here was
conducted for all 2429 reefs to create DHW projections spanning from 2025 to 2075. Code used
to calculate DHW data and perform analyses involved in this investigation can be found at
Grier and Iwanaga [-@grier2025].

@@

@@

Connectivity data from 24 days in total were used in this investigation. These data span six
years from 2015 to 2020. The years 2015, 2016, 2018 and 2019 are each represented by
connectivity matrices covering spawning periods over three days in each year (12 days in
total), while 2017 and 2020 are each represented by six spawning days each, coming to a
total of 24 days of connectivity matrices. Indicative connectivity strengths are implemented
in ADRIA-CoralBlox as the mean of all available data for each location (i.e., mean of all 24
spawning days). While the connectivity strengths do not vary from year to year within each
simulation, they do vary between reefs which are of varying sizes, providing sufficient
heterogeneity to capture spatial patterns in larval dispersal and settlement dynamics. The
spatial variability also enables investigations into the influence of connectivity strength
and its role in driving recovery processes.

The ratio of source-to-sink connectivity ($O:I$) for a target reef represents the strength
of the reef's outgoing larval flow relative to it's incoming larval flow (@eq-so-to-si). If
a reef has no incoming connections, the $O:I$ equals the $S_O/N_O$ value. Incoming
connections in this calculation, do not directly represent the larval supply gained as the
population from which connections are coming from is not represented. A source-to-sink ratio
greater than 1 (a value greater than 0 on a log scale) indicates a reef has greater outgoing
than incoming larval flow.

@@

ADRIA-CoralBlox was run for 4096 scenarios for each of the five GCMs (20,480 total model
runs). Here, a scenario represents a unique combination of assumed environmental (namely one
of the five DHW trajectories) and ecological factors including coral growth and mortality
rates, coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol'
sampling method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a random-Quasi
Monte Carlo approach which balances scenario exploration with improved convergence
properties compared to pseudo-random sampling. A detailed comparative analysis of the
sampling methods can be found in [@kucherenko2015]. For the purpose of scenario assessment,
the resulting time series for each model run were scaled by the total area of the reef to
represent cover as a percentage of the total estimated reef area. These time series of coral
cover were then clustered based on their temporal behaviour.

@@

The time series clustering was applied over the time frame between 2025 and 2075. Spatially,
the analyses were conducted for each GBR management area and at the scale of each reefal
bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were also performed at the
whole-of-GBR scale but were found to lead to identical conclusions about characteristics
influencing cluster cover compared to management area and bioregion scale analyses, and so
are not shown here. To compare time series clustering results across GCMs, Management area
scale clustering results are used to preserve coarse scale latitudinal differences between
regions, while maintaining ease of comparison and visualisation. These comparisons include
quantification of differences between low and high cover clusters across management regions
under each GCM using the CID between cluster median time series. To assess how different
reef characteristics influence time series clustering at a finer spatial scale, bioregion
results are used to capture patterns occurring within individual clusters of reefs. Results
of all analyses at all clustering scales can be found at Grier and Iwanaga [-@grier2025] for
those interested.

Reefs in each time series cluster were compared to identify characteristics that differ
between Low, Medium and High categories. These characteristics include reef median depth,
mean time series DHW, mean time series DHW tolerance level **\[TI: Ben, still correct? Still
included somewhere?\]**, Pearson’s correlation between total coral cover and inverse DHW
time series, proportion of coral habitable area occupied by corals in the first timestep,
ratio of source-to-sink larval connectivity and total strength of all larval connections
between reefs.

@@

Using median coral cover time series for each reef we calculated the number of years each
reef maintains coral cover above a given percent threshold. The threshold represents the
level of live coral cover necessary to maintain a positive carbonate budget (denoted here as
$\theta_{0}$ to indicate the budget equilibrium). Greater number of years above $\theta_{0}$
therefore suggests greater resilience and opportunity for recovery.

Recent analyses suggest a general threshold of \~17% live coral cover in the case of the GBR
[@desbiens2021]. Given in situ carbonate budgets are influenced by (1) a range of processes;
(2) the variety of approaches that could be used to measure those processes; and (3) the
uncertainty surrounding these, it would be unreasonable to assume all reefs adhere to a
single precise $\theta_{0}$ value. Concurrently, it is difficult to ascertain a suitable
$\theta_{0}$ value for each individual reef.

@@

**\[BG: I could combine these two plots into a single larger figure maybe, since it's quite
easy to see the contrasting depth vs connectivity patterns with colours even when zoomed
out.\]**

@@

Our findings suggest that reefs in ‘high’ cover clusters do not have greater thermal
tolerance than reefs in lower coral cover clusters. This is in contrast with literature
suggesting corals have the capacity to adapt to increasing climate change pressures
[@humanes2022; @lachs2024; @lachs2023], including the investigation by Bairos-Novak et al.
[-@bairos-novak2021] that revealed high heritability of bleaching and survival response
traits, used to inform thermal tolerance trait distributions in this investigation. The low
importance of thermal tolerance on cover outcomes may be due to reefs experiencing high
levels of thermal stress early in projected time series. High - early time series thermal
stress events may prevent the intergenerational adaptation of corals to thermal stress and
prevent thermal tolerance from reaching levels greater than 9°C-weeks. **\[TI note: This
paragraph needs to be revisited\]**

@@

Additionally, it is currently not possible to obtain highly accurate and precise estimates
of reef depth at the GBR-scale. Data adopted for the given assessment are said to be
accurate up to 15m from Mean Sea Level [@GBRMPA2020], however there are many sources of
uncertainty in the depth estimates used for this study. Satellite-derived estimates are
subject to different bottom types (rock, sand, seagrass, etc) and how these affect
reflectance, the water penetration depth of satellite sensors, influence of local conditions
(e.g., turbidity and wave activity), among many other factors. The median depths for each
indicative reef area were extracted from bathymetry data using zonal statistics and depth
values beyond the stated 15m accuracy threshold [@GBRMPA2020] were included. The resulting
depth estimates are therefore highly uncertain and are only adopted as a useful indication
for the purpose of the presented analysis. They are not intended to be an accurate and
precise report of reef depths across the GBR.

@@

CoralBlox model incorporates linear extension (growth) and background mortality rates that
are dependent on the size and functional group of a coral colony. Environmental disturbances
and reproduction are not directly considered within CoralBlox. Instead, these factors are
considered by making changes to the distribution of colonies within CoralBlox timesteps
using the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform [@iwanaga2023].
Within ADRIA, we incorporate coral life history and environmental events in the following
order: coral colony growth; coral reproduction; larval connectivity and settlement; and
coral bleaching due to heat stress. As ADRIA and CoralBlox are coupled together to represent
the system, we refer to both as ADRIA-CoralBlox.

@@

**Bleaching mortality** **\[BG: Has this section been completed?\]** In ADRIA we track the
thermal tolerance of each functional group and size class as a truncated normal distribution
that is updated in each timestep following growth and recruitment. This distribution
determines the proportion of the population on a reef that is susceptible to bleaching (i.e.
has a thermal threshold below the experienced DHW value). Starting distributions for thermal
tolerance levels are informed by Bairos-Novak et al. [-@bairos-novak2021].

@@

**\[TI: To be revisited\]**

The CID corrects the Euclidean distance between pairs of time series (which accounts for
magnitude differences) with the complexity of each time series (obtained through a measure
of their variability) to ensure reefs are clustered based on their underlying temporal
dynamics rather than simply their average coral cover.

The process creates a distance matrix of n \* n dimensions (where n is the number of time
series in the group) that is clustered using a k-medoids clustering algorithm [cf.
@steinmann2020]. The three reef clusters are labelled with qualitative-ordered categories
from Low, Medium and High based on their median cover relative to each other cluster over
the 2030-2060 period.The selected time frame prevents the classification from being overly
influenced by any steep declines in the initial years of simulations, and avoids the
post-2060 time period where reefs are projected to be under severe stress and in a declined
state [@bozec2025].

@@

![Mean time series DHW tolerance value, from 2035 to 2075, for reefs within each bioregion,
separated into clusters assigned by time series clustering. ‘Low’ clusters (green) are
clusters with the lowest median area-relative cover between 2030 and 2060 within each
bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative
cover between 2030 and 2060 within each bioregion. ‘High’ clusters (blue) are clusters with
the highest median area-relative cover between 2030 and 2060 within each bioregion. Solid
lines represent the median DHW tolerance value for each cluster and bands represent the
distribution of DHW-tolerance levels within each cluster. Displays results for ACCESS-ESM1-5
GCM.](../figs/ACCESS-ESM1-5/bioregion/dhw_tolerance_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwtol
fig-align="center" width="14.82cm"}

>>>>

Depth characteristics for each reef were extracted using 10m resolution bathymetry data made
available in GBRMPA [-@GBRMPA2020]. For each reef, bathymetry data were spatially matched
with the 3806 reef outlines through geometric intersections. All depth values were converted
to positive values (depths below sea level). The median depth was used to provide an
indicative depth value for each reef. Code, and the resulting depth estimates, are available
at **\[cite Canonical Reefs repo\]**.

While the bathymetry data adopted for the given assessment are said to be accurate up to 15m
from Mean Sea Level [@GBRMPA2020], there are many sources of uncertainty in the depth
estimates used for this study. These are outlined further in the Discussion. Notably, the
adopted depth estimates were derived from bathymetries based on satellite imagery. The
bathymetry data can be limited in their alignment with the established GBRMPA reef outlines
used to characterise reef locations and their areal extent. A Quality Assessment and Control
process identified reefs with potential data limitations: reefs with no overlapping
bathymetry data, reefs with minimum depth values above sea level (indicating potential data
artifacts), and reefs where bathymetry pixels covered less than 5% of the reef polygon area.
This process removed 842 reefs, or 22%, from the initial set of 3806 reefs. Furthermore,
reefs that are not associated with GBRMPA Management areas [@GBRMPA2007] or reefal
bioregions [@GBRMPA2024] were also removed from the investigation. Bioregions which
contained fewer than three reefs after these processes were also excluded. As a result, 1377
reefs were excluded (36%), leaving a total of 2429 remaining for the purpose of this
investigation.

@@

These GCMs were selected based on evaluation conducted by Grose et al., [-@grose2023],
indicating their ability to recreate historical atmospheric temperature and precipitation
conditions in the Australasian region. They span a range of Equilibrium Climate Sensitivity
(ECS) values within the Intergovernmental Panel on Climate Change (IPCC) assessed likely
range. The IPCC defines ECS as the long term warming caused by a doubling of carbon dioxide
above its pre-industrial concentration [@forster2021]. While this assessment does not
consider oceanic data, Grose et al. [-@grose2023] provide a region-specific selection
mechanism which we use as an indicator of general performance of GCMs for the region. An
overview of the selected GCMs are provided in @tbl-gcms. Note here that Grose et al.,
[-@grose2023] assessed GFDL-ESM4, whereas Dixon et al., [-@dixon2022; -@dixon2023] provides
data based on GFDL-CM4. This substitution is assumed appropriate given they share the same
fundamental representation of the physical climate system.

Projections using Shared Socio-economic Pathway (SSP) 2-4.5 was used to represent the
plausible emission pathway. It represents a moderate emissions scenario with continued CO2
emissions comparable to present day until 2050, after which emissions gradually decline
[@hausfather2020; IPCC -@climate2023]. The selected GCMs represent 2.5°C to 4.66°C of global
atmospheric warming in the 2081-2100 period relative to 1850-1900. The use of these GCMs
therefore allow for a range of possible climate futures to be explored, capturing the
uncertainty in future responses to CO2 increases (@sfig-ecs; @fig-methods-dhws; [@grose2023;
@shepherd2018]) and enables the assessment of model projected coral reef behaviour over a
broad range of independently derived climate futures [@abramowitz2019].

@@

Degree Heating Weeks (DHWs) were derived using the "HotSpot SST Anomaly" approach
[@skirving2020]. The "HotSpot SST Anomaly" represents thermal stress experienced by coral
reefs and is calculated as the difference between daily SST values and the maximum of a
pixel's monthly mean climatology across all twelve months (i.e., the maximum monthly mean
SST). This approach uses the annual maximum temperature as the baseline because corals are
adapted to withstand their local warmest conditions under normal seasonal cycles. When
temperatures exceed this seasonal maximum, corals experience thermal stress that can lead to
bleaching. Only positive anomalies, greater than 1°C, are considered relevant for coral
stress, as temperatures below the seasonal maximum do not contribute to bleaching risk.

DHW values on a per-pixel basis at 1km^2^ resolution were then calculated as the 12-week
rolling sum of HotSpot SST Anomaly values that are greater than 1°C, expressed in units of
°C-weeks. The 12-week accumulation period reflects the cumulative nature of coral thermal
stress. Brief temperature spikes may be tolerable, but sustained or repeated exposure to
elevated temperatures causes progressively greater physiological stress and bleaching
severity. DHW values integrate both the magnitude and duration of thermal stress: a moderate
anomaly (e.g., 1°C) sustained for 12 weeks produces the same DHW value as a severe anomaly
(e.g., 3°C) lasting 4 weeks, though the biological impacts may differ. DHW estimates for
each indicative reef (see @GBRMPA2022) were then determined by taking the mean DHW value for
all pixels touching each reef outline. The process described here was conducted for all 2429
reefs to create DHW projections spanning from 2025 to 2075. Code used to calculate DHW data
and perform analyses involved in this investigation can be found at Grier and Iwanaga
[-@grier2025].

@@

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a random-Quasi Monte Carlo
approach which balances scenario exploration with improved convergence properties compared
to pseudo-random sampling. A detailed comparative analysis of the sampling methods can be
found in [@kucherenko2015]. For the purpose of scenario assessment, the resulting coral
cover time series for each model run were scaled by the total area of each reef to represent
cover as a percentage of the total estimated reef area. The scenario-median timeseries for
each reef was used for further analyses. These time series of coral cover were then
clustered based on their temporal behaviour.

@@

Connectivity data from 24 days in total were used in this investigation to estimate coral
larval flows. These data span six years from 2015 to 2020. The years 2015, 2016, 2018 and
2019 are each represented by connectivity matrices covering spawning periods over three days
in each year (12 days in total), while 2017 and 2020 are each represented by six spawning
days each, coming to a total of 24 days of connectivity matrices. Indicative connectivity
strengths are implemented in ADRIA-CoralBlox as the mean of all available data for each
location (i.e., mean of all 24 spawning days). While the connectivity strengths do not vary
from year to year within each simulation, they do vary between reefs which are of varying
sizes, providing sufficient heterogeneity to capture spatial patterns in larval dispersal
and settlement dynamics. The spatial variability also enables investigations into the
influence of connectivity strength and its role in driving recovery processes.

To assess the impact of reef connectivity on modelled outcomes we summarise inter-reef
connections using the total strength of connections and the ratio of source-to-sink
connectivity for each reef. The ratio of source-to-sink connectivity ($O:I$) for a target
reef represents the strength of the reef's outgoing larval flow relative to it's incoming
larval flow (@eq-so-to-si). If a reef has no incoming connections, the $O:I$ equals the
$S_O/N_O$ value. Incoming connections in this calculation, do not directly represent the
larval supply gained as the population from which connections are coming from is not
represented. A source-to-sink ratio greater than 1 (a value greater than 0 on a log scale)
indicates a reef has greater outgoing than incoming larval flow.

@@

@@

The time series clustering was applied over the time frame between 2025 and 2075. Spatially,
the clustering analyses were conducted for each GBR management area and at the scale of each
reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were also performed at the
whole-of-GBR scale but were found to lead to identical conclusions about characteristics
influencing cluster cover compared to management area and bioregion scale analyses, and so
are not shown here. To compare time series clustering results across GCMs, management area
scale clustering results are used to preserve coarse scale latitudinal differences between
regions, while maintaining ease of comparison and visualisation. These comparisons include
quantification of differences between low and high cover clusters across management regions
under each GCM using the CID between cluster median time series. To assess how different
reef characteristics influence time series clustering at a finer spatial scale, bioregion
results are used to capture patterns occurring within individual groups of reefs. Results of
all analyses at all clustering scales can be found at Grier and Iwanaga [-@grier2025] for
those interested.

Reefs in each time series cluster were compared to identify characteristics that differ
between Low, Medium and High categories. These characteristics include reef median depth,
mean time series DHW, Pearson’s correlation between total coral cover and inverse DHW time
series, proportion of coral habitable area occupied by corals in the first timestep, ratio
of source-to-sink larval connectivity and total strength of all larval connections between
reefs.

@@

Using coral cover time series for each reef we calculated the number of years each reef
maintains coral cover above a given percent threshold. The threshold represents the level of
live coral cover necessary to maintain a positive carbonate budget (denoted here as
$\theta_{0}$ to indicate the budget equilibrium). Greater number of years above $\theta_{0}$
therefore suggests greater resilience and opportunity for recovery.

Recent analyses suggest a general threshold of \~17% live coral cover in the case of the
mid-shelf and offshore GBR reefs [@desbiens2021]. Given in situ carbonate budgets are
influenced by (1) a range of processes; (2) the variety of approaches that could be used to
measure those processes; and (3) the uncertainty surrounding these, it would be unreasonable
to assume all reefs adhere to a single precise $\theta_{0}$ value. Concurrently, it is
difficult to ascertain a suitable $\theta_{0}$ value for each individual reef.

@@

@@

@@

Additionally, it is currently not possible to obtain highly accurate and precise estimates
of reef depth at the GBR-scale. Data adopted for the given assessment are said to be
accurate up to 15m from Mean Sea Level [@GBRMPA2020; @li2021], however there are many
sources of uncertainty in the depth estimates used for this study. Satellite-derived
estimates are subject to different bottom types (rock, sand, seagrass, etc) and how these
affect reflectance, the water penetration depth of satellite sensors, influence of local
conditions (e.g., turbidity and wave activity), among many other factors. The median depths
for each indicative reef area were extracted from bathymetry data using zonal statistics and
depth values beyond the stated 15m accuracy threshold [@GBRMPA2020] were included. The
resulting depth estimates are therefore highly uncertain and are only adopted as a useful
indication for the purpose of the presented analysis. They are not intended to be an
accurate and precise report of reef depths across the GBR.

@@

The CoralBlox model incorporates linear extension (growth) and background mortality rates
that are dependent on the size and functional group of a coral colony. Environmental
disturbances and reproduction are not directly considered within CoralBlox. Instead, these
factors are considered by making changes to the distribution of colonies within CoralBlox
timesteps using the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform
[@iwanaga2023]. Within ADRIA, we incorporate coral life history and environmental events in
the following order: coral colony growth; coral reproduction; larval connectivity and
settlement; and coral bleaching due to heat stress. As ADRIA and CoralBlox are coupled
together to represent the system, we refer to both as ADRIA-CoralBlox.

@@

In ADRIA we track the thermal tolerance of each functional group and size class as a
truncated normal distribution that is updated in each timestep following growth and
recruitment. This distribution determines the proportion of the population on a reef that is
susceptible to bleaching (i.e. has a thermal threshold below the experienced DHW value).
Starting distributions for thermal tolerance levels are informed by Bairos-Novak et al.
[-@bairos-novak2021].

@@

The CID timeseries clustering metric corrects the Euclidean distance between pairs of time
series (which accounts for magnitude differences) with the complexity of each time series
(obtained through a measure of their variability) to ensure reefs are clustered based on
their underlying temporal dynamics rather than simply their average coral cover. The process
creates a distance matrix of n \* n dimensions (where n is the number of time series in the
group) that is clustered using a k-medoids clustering algorithm [cf. @steinmann2020]. The
three reef clusters are labelled with qualitative-ordered categories from Low, Medium and
High based on their median cover relative to each other cluster over the 2030-2060 period.
The selected time frame prevents the classification from being overly influenced by any
steep declines in the initial years of simulations, and avoids the post-2060 time period
where reefs are projected to be under severe stress and in a declined state [@bozec2025].

@@
